---
title: "DATA 643 Prep: Chapter 3, Data Mining Techniques"
output:
  html_document:
    theme: lumen

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Text

Building a Recommendation System with R
by Suresh K. Gorakala (Author), Michele Usuelli (Author) 
https://www.amazon.com/dp/B012O8S1YM/ref=dp-kindle-redirect?_encoding=UTF8&btkr=1

# The Recommender Lab Package

```{r, message=FALSE}
library(recommenderlab)
library(ggplot2)
data_package <- data(package = "recommenderlab")
data_package$results[,'Item']
```

## Datasets

```{r}
data("MovieLense")
MovieLense
class(MovieLense)
methods(class = class(MovieLense))
```

## Object Storage

```{r}
object.size(MovieLense) #since sparse matrix, more efficient way to store
object.size(as(MovieLense, "matrix"))
object.size(as(MovieLense, "matrix")) / object.size(MovieLense) #how much smaller?
```

## Similarity Matrix

```{r}
similarity_users <- similarity(MovieLense[1:5,], method="cosine", which="users")
as.matrix(similarity_users)
image(as.matrix(similarity_users), main="User similarity")
similarity_items <- similarity(MovieLense[, 1:5], method="cosine", which="items")
as.matrix(similarity_items)
image(as.matrix(similarity_items), main="Item similarity")
```

## Recommendation Models

```{r}
recommender_models <- recommenderRegistry$get_entries(dataType = "realRatingMatrix")
names(recommender_models)
lapply(recommender_models, "[[", "description")
recommender_models$IBCF_realRatingMatrix$parameters
```

# Data Exploration

```{r}
dim(MovieLense) # users, movies
slotNames(MovieLense)
class(MovieLense@data)
```

## Exploring the values of the rating

```{r}
vector_ratings <- as.vector(MovieLense@data)
unique(vector_ratings)
vector_ratings <- factor(vector_ratings[vector_ratings != 0])
table_ratings <- table(vector_ratings)
table_ratings
qplot(vector_ratings) + ggtitle("Distribution of Ratings")
```

## Exploring which movies have been viewed

```{r}
views_per_movie <- colCounts(MovieLense)
table_views <- data.frame(
    movie = names(views_per_movie),
    views = views_per_movie
)
table_views <- table_views[order(table_views$views, decreasing=T),]
ggplot(table_views[1:6, ], aes(x = movie, y = views)) + 
    geom_bar(stat="identity") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ggtitle("Number of Views of the Top Movies")
```

## Exploring the average ratings

```{r, message=F, warning=F}
average_ratings <- colMeans(MovieLense)
qplot(average_ratings) + stat_bin(binwidth = 0.1) + 
    ggtitle("Distribution of the Avg Movie Rating")
average_ratings_relevant <- average_ratings[views_per_movie > 100]
qplot(average_ratings_relevant) + stat_bin(binwidth = 0.1) + 
    ggtitle("Distribution of the Relevant Avg Movie Rating")
```

## Visualizing the Matrix

```{r}
image(MovieLense, main="Heatmap of the Rating Matrix")
image(MovieLense[1:10, 1:15], main="Heatmap of the first rows and columns")

## get most relevant users (seen most movies / most seen movies)
min_n_movies <- quantile(rowCounts(MovieLense), 0.99)
min_n_users <- quantile(colCounts(MovieLense), 0.99)
image(MovieLense[rowCounts(MovieLense) > min_n_movies, 
                 colCounts(MovieLense) > min_n_users], 
      main="Heatmap of the top users and movies")
```

# Data Preparation