---
title: "DATA 643 Prep: Chapter 5, Building Your Own Recommendation Engine"
output:
  html_document:
    theme: lumen

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Text

Building a Recommendation System with R
by Suresh K. Gorakala (Author), Michele Usuelli (Author) 
https://www.amazon.com/dp/B012O8S1YM/ref=dp-kindle-redirect?_encoding=UTF8&btkr=1

# Env Prep

```{r, message=F, warning=F}
library("data.table")
library("ggplot2")
library("recommenderlab")
library("countrycode")
```

# Data Preparation {.tabset .tabset-pills} 

## Importing the data

```{r}
file_in <- "anonymous-msweb.test.txt"
table_in <- read.csv(file_in, header=F)
head(table_in)
table_users <- table_in[, 1:2]
table_users <- data.table(table_users)
setnames(table_users, 1:2, c("category", "value"))
table_users <- table_users[category %in% c("C", "V")]
head(table_users)
```

## Defining a rating matrix

```{r}
table_users[, chunk_user := cumsum(category =="C")]
head(table_users)
table_long <- table_users[, list(user = value[1], item = value[-1]), by = "chunk_user"]
head(table_long)
table_long[, value := 1]
table_wide <- reshape(data = table_long,
                      direction = "wide", 
                      idvar = "user", 
                      timevar = "item", 
                      v.names = "value")
head(table_wide[, 1:5, with=F])
vector_users <- table_wide[, user]
table_wide <- table_wide[, user := NULL]
table_wide <- table_wide[, chunk_user := NULL]

setnames(x = table_wide,
         old=names(table_wide),
         new=substring(names(table_wide), 7))
matrix_wide <- as.matrix(table_wide)
rownames(matrix_wide) <- vector_users

## coerce to binary
matrix_wide[is.na(matrix_wide)] <- 0
ratings_matrix <- as(matrix_wide, "binaryRatingMatrix")
ratings_matrix
```

## Heatmap

```{r}
image(ratings_matrix[1:50, 1:50], main="Binary Rating Matrix")
```

## Distrubtion of Users

```{r}
n_users <- colCounts(ratings_matrix)
qplot(n_users[n_users > 100]) + stat_bin(bindwidth = 10) + ggtitle("Distribution of the number of users")
```

## Remove Items 

```{r}
# too few purchases
ratings_matrix <- ratings_matrix[, colCounts(ratings_matrix) >= 5]
ratings_matrix <- ratings_matrix[rowCounts(ratings_matrix) >= 5, ]
ratings_matrix
```

# Exacting Item Attributes

```{r}
table_in <- data.table(table_in)
table_items <- table_in[V1 == "A"]
head(table_items)
table_items <- table_items[, c(2, 4, 5), with = F]
setnames(table_items, 1:3, c("id", "description", "url"))
table_items <- table_items[order(id)]
head(table_items)
table_items <- table_items[, category := "product"]

name_countries <- c(countrycode_data$country.name.en, "Taiwan", "UK", "Russia",
                    "Venezuela", "Slovenija", "Caribbean", "Netherlands (Holland)",
                    "Europe", "Central America", "MS North Africa")
table_items <- table_items[description %in% name_countries, category := "region"]
table_items <- table_items[grepl("Region", description), category := "region"]
head(table_items)
table_items[, list(n_items= .N), by=category]
```

# Building the Model {.tabset .tabset-pills} 

```{r}
which_train <- sample(x = c(TRUE, FALSE), 
                      size = nrow(ratings_matrix), 
                      replace = TRUE, 
                      prob = c(0.8, 0.2))
recc_data_train <- ratings_matrix[which_train, ]
recc_data_test <- ratings_matrix[!which_train,]

recc_model <- Recommender(data = recc_data_train, 
                          method = "IBCF", 
                          parameter = list(method = "Jaccard"))
image(recc_model@model$sim)
```

## Combine Distance matrix with descriptions

```{r}
dist_ratings <- as(recc_model@model$sim, "matrix")
dist_category <- table_items[, 1 - dist(category == "product")]
dist_category <- as(dist_category, "matrix")
dim(dist_category); dim(dist_ratings) # different sizes, because model only rated those purchased
```

## Match itemnames

```{r}
rownames(dist_category) <- table_items[, id]
colnames(dist_category) <- table_items[, id]
vector_items <- rownames(dist_ratings)
dist_category <- dist_category[vector_items, vector_items] #subset
```

## Show sorted map of ratings Dist

```{r}
image(dist_category)
```

## Combine Tables | add to model

```{r}
weight_category <- 0.25
dist_tot <- dist_category * weight_category + dist_ratings * (1 - weight_category)
recc_model@model$sim <- as(dist_tot, "dgCMatrix")
image(dist_tot)
```

## Predict

```{r}
n_recommended <- 10
recc_predicted <- predict(object = recc_model, 
                          newdata = recc_data_test, 
                          n = n_recommended)
table_labels <- data.frame(id = recc_predicted@itemLabels)
table_labels <- merge(table_labels, table_items, by="id", 
                      all.x = TRUE, all.y = FALSE, sort = FALSE)
descriptions <- as(table_labels$description, "character")
head(table_labels)
```

## Extract User

```{r}
recc_user_1 <- recc_predicted@items[[1]]
items_user_1 <- descriptions[recc_user_1]
#head(items_user_1)
recc_matrix <- sapply(recc_predicted@items, function(x){
    recommended <- descriptions[x]
    c(recommended, rep("", n_recommended - length(recommended)))
})
dim(recc_matrix)
head(recc_matrix[,1:3])
```