---
title: "DATA 643 Prep: Chapter 3, Data Mining Techniques"
output:
  html_document:
    theme: lumen

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Text

Building a Recommendation System with R
by Suresh K. Gorakala (Author), Michele Usuelli (Author) 
https://www.amazon.com/dp/B012O8S1YM/ref=dp-kindle-redirect?_encoding=UTF8&btkr=1

# The Recommender Lab Package

```{r, message=FALSE}
library(recommenderlab)
library(ggplot2)
data_package <- data(package = "recommenderlab")
data_package$results[,'Item']
```

## Datasets

```{r}
data("MovieLense")
MovieLense
class(MovieLense)
methods(class = class(MovieLense))
```

## Object Storage

```{r}
object.size(MovieLense) #since sparse matrix, more efficient way to store
object.size(as(MovieLense, "matrix"))
object.size(as(MovieLense, "matrix")) / object.size(MovieLense) #how much smaller?
```

## Similarity Matrix

```{r}
similarity_users <- similarity(MovieLense[1:5,], method="cosine", which="users")
as.matrix(similarity_users)
image(as.matrix(similarity_users), main="User similarity")
similarity_items <- similarity(MovieLense[, 1:5], method="cosine", which="items")
as.matrix(similarity_items)
image(as.matrix(similarity_items), main="Item similarity")
```

## Recommendation Models

```{r}
recommender_models <- recommenderRegistry$get_entries(dataType = "realRatingMatrix")
names(recommender_models)
lapply(recommender_models, "[[", "description")
recommender_models$IBCF_realRatingMatrix$parameters
```

# Data Exploration

```{r}
dim(MovieLense) # users, movies
slotNames(MovieLense)
class(MovieLense@data)
```

## Exploring the values of the rating

```{r}
vector_ratings <- as.vector(MovieLense@data)
unique(vector_ratings)
vector_ratings <- factor(vector_ratings[vector_ratings != 0])
table_ratings <- table(vector_ratings)
table_ratings
qplot(vector_ratings) + ggtitle("Distribution of Ratings")
```

## Exploring which movies have been viewed

```{r}
views_per_movie <- colCounts(MovieLense)
table_views <- data.frame(
    movie = names(views_per_movie),
    views = views_per_movie
)
table_views <- table_views[order(table_views$views, decreasing=T),]
ggplot(table_views[1:6, ], aes(x = movie, y = views)) + 
    geom_bar(stat="identity") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ggtitle("Number of Views of the Top Movies")
```

## Exploring the average ratings

```{r, message=F, warning=F}
average_ratings <- colMeans(MovieLense)
qplot(average_ratings) + stat_bin(binwidth = 0.1) + 
    ggtitle("Distribution of the Avg Movie Rating")
average_ratings_relevant <- average_ratings[views_per_movie > 100]
qplot(average_ratings_relevant) + stat_bin(binwidth = 0.1) + 
    ggtitle("Distribution of the Relevant Avg Movie Rating")
```

## Visualizing the Matrix

```{r}
image(MovieLense, main="Heatmap of the Rating Matrix")
image(MovieLense[1:10, 1:15], main="Heatmap of the first rows and columns")

## get most relevant users (seen most movies / most seen movies)
min_n_movies <- quantile(rowCounts(MovieLense), 0.99)
min_n_users <- quantile(colCounts(MovieLense), 0.99)
image(MovieLense[rowCounts(MovieLense) > min_n_movies, 
                 colCounts(MovieLense) > min_n_users], 
      main="Heatmap of the top users and movies")
```

# Data Preparation

## Select Most Relevant Data

```{r}
ratings_movies <- MovieLense[rowCounts(MovieLense) > 50, colCounts(MovieLense) > 100]
ratings_movies
```

## Explore Most Relevant Data

```{r}
# top 2%
min_movies <- quantile(rowCounts(ratings_movies), 0.98)
min_users <- quantile(colCounts(ratings_movies), 0.98)
image(ratings_movies[rowCounts(ratings_movies) > min_movies, 
                 colCounts(ratings_movies) > min_users], 
      main="Heatmap of the top users and movies")

average_ratings_per_user <- rowMeans(ratings_movies)
qplot(average_ratings_per_user) + stat_bin(binwidth = 0.1) + 
    ggtitle("Distribution of the Avg Rating Per User")
```

## Normalizing the Data

```{r}
ratings_movies_norm <- normalize(ratings_movies)
sum(rowMeans(ratings_movies_norm) > 0.00001)
image(ratings_movies_norm[rowCounts(ratings_movies_norm) > min_movies, 
                 colCounts(ratings_movies_norm) > min_users], 
      main="Heatmap of the top users and movies")
```

## Binarizing the Data

```{r}
ratings_movies_watched <- binarize(ratings_movies, minRating=1)
min_movies_binary <- quantile(rowCounts(ratings_movies), 0.95)
min_users_binary <- quantile(colCounts(ratings_movies), 0.95)
image(ratings_movies_watched[rowCounts(ratings_movies) > min_movies_binary, 
                             colCounts(ratings_movies) > min_users_binary], 
      main = "Heatmap of the top users and movies")
ratings_movies_good <- binarize(ratings_movies, minRating=3)
image(ratings_movies_good[rowCounts(ratings_movies) > min_movies_binary, 
                             colCounts(ratings_movies) > min_users_binary], 
      main = "Heatmap of the top users and movies, rating > 3")
```

# Item based Collaborative Filtering

## Define training and test 

```{r}
which_train <- sample(x = c(TRUE, FALSE), size = nrow(ratings_movies), 
                      replace = TRUE, prob = c(0.8, 0.2))
recc_data_train <- ratings_movies[which_train, ]
recc_data_test <- ratings_movies[-which_train, ]
```

## Building the Recommendation Model

```{r}
recc_model <- Recommender(data = recc_data_train, 
                          method = "IBCF", 
                          parameter = list(k = 30))
recc_model
```

## Exploring the Recommender Model

```{r}
model_details <- getModel(recc_model)
model_details$description
model_details$k
class(model_details$sim)
n_items_top <- 20
image(model_details$sim[1:n_items_top, 1:n_items_top], main = "Heatmap of First rows and columns")
model_details$k
row_sums <- rowSums(model_details$sim > 0)
table(row_sums)
col_sums <- colSums(model_details$sim > 0)
qplot(col_sums) + stat_bin(binwidth=1) + ggtitle("Distribution of the Column Count")
which_max <- order(col_sums, decreasing=TRUE)[1:6]
rownames(model_details$sim)[which_max]
```

## Applying the model

```{r}
n_recommended <- 6
recc_predicted <- predict(object = recc_model, newdata = recc_data_test, n = n_recommended)
recc_predicted
slotNames(recc_predicted)

## extract rec for user
recc_user_1 <- recc_predicted@items[[1]]
movies_user_1 <- recc_predicted@itemLabels[recc_user_1]
movies_user_1

## matrix w/ recc
recc_matrix <- sapply(recc_predicted@items, function(x){colnames(ratings_movies)[x]})
dim(recc_matrix)
recc_matrix[, 1:4]

# dist
number_of_items <- factor(table(recc_matrix))
qplot(number_of_items) + ggtitle("Distribution of the number of items for IBCF")

# which movies are recc most
number_of_items_sorted <- sort(number_of_items, decreasing = T)
number_of_items_top <- head(number_of_items_sorted, n = 4)
table_top <- data.frame(names(number_of_items_top), number_of_items_top)
table_top
```

## Collaborative filtering on binary data

```{r}
ratings_movies_watched <- binarize(ratings_movies, minRating=1)
qplot(rowSums(ratings_movies_watched)) + stat_bin(binwidth = 10) + geom_vline(xintercept = mean(rowSums(ratings_movies_watched)), col="red", linetype ="dashed") + ggtitle("Dist of Movies by User")
```

## Item-based collaborative filtering on binary data

```{r}
recc_model <- Recommender(data = recc_data_train, method="IBCF", parameter=list(method="Jaccard")) #cosine sim or pearson ineffective on binary, just used jaccard
model_details <- getModel(recc_model)
n_recommended <- 6
recc_predicted <- predict(object = recc_model, newdata = recc_data_test, n=n_recommended)
recc_matrix <- sapply(recc_predicted@items, function(x){colnames(ratings_movies)[x]})
recc_matrix[,1:4]
```

## User-based Collaborative filtering on binary data

```{r}
recc_model <- Recommender(data = recc_data_train, method="UBCF", parameter=list(method="Jaccard"))
model_details <- getModel(recc_model)
n_recommended <- 6
recc_predicted <- predict(object = recc_model, newdata = recc_data_test, n=n_recommended)
recc_matrix <- sapply(recc_predicted@items, function(x){colnames(ratings_movies)[x]})
recc_matrix[,1:4]
```